Designing an effective intervention of motivational interview using sequence classification methods

### 2017-02-27 paper/AMIA/main.tex
update introduction section of DMC paper for AMIA 2017

\title{Leverage machine learning and textual data mining methods for the classification of clinical interview fragments and their sequences}

@@

\noindent{\bf Abstract}
\textit{This study examines the effectiveness of state-of-the-art supervised machine learning methods for the task of automatic annotation of fragments of clinical interviews in conjunction with sequential analysis of that annotation. We used a collection of motivational interview transcripts consisting of 11,353 utterances, which were manually annotated by two human coders as a golden standard, and experimented with state-of-art classifiers, including Na\"{i}ve Bayes, J48 Decision Tree, Support Vector Machine (SVM), Random Forest (RF), AdaBoost, DiscLDA, Conditional Random Fields (CRF) and Convolutional Neural Network (CNN) in conjunction with lexical, contextual (label of the previous utterance) and semantic (distribution of words in the utterance across the Linguistic Inquiry and Word Count dictionaries) features. We found out that, when the number of classes is large, the performance of the CNN and CRFs is inferior to SVM. Among all classifiers using only lexical features, SVM achieved the highest classification accuracy of 70.8\%, 61\% and 53.7\%, when interview transcripts were annotated with the codebooks consisting of 17, 20 and 41 codes, respectively. Using contextual and semantic features, as well as their combination, in addition to lexical ones improved the accuracy of SVM for annotation of utterances in motivational interview transcripts with a codebook consisting of 17 classes to 71.5\%, 74.2\% and 75.1\%, respectively. We also used first order markov model that achieved 68.03\% accuracy for the classification of sequence of behavioral codes extracted from the above dataset. Therefore, our results demonstrate the potential of using machine learning  and textual data mining methods in conjunction with semantic and contextual features for automatic annotation of clinical interview transcripts  and their sequences with near-human accuracy.}

\section*{Introduction}
In this study, our analysis have two directions: automatic annotation and sequential data analysis. Annotation (or labeling) of fragments of clinical text with the categories (or labels, codes) from a predefined codebook is an integral part of qualitative research. It can also be viewed as classification of textual fragments into a predefined number of categories (classes). Textual annotation has been traditionally performed manually by trained coders, which is a tedious, costly and time-consuming process. Furthermore, manual annotation increases the likelihood of errors due to coder fatigue and bias associated with human subjectivity. To automate tedious cognitive tasks such as classification, supervised machine learning methods have been recently proposed. Although these methods have been shown to be successful at binary (two-class) classification \citep{40,41} (e.g. classifying textual fragments as neutral or opinionated), we are not aware of any prior studies reported in the literature that evaluate their performance for textual classification tasks involving large number of classes. Such tasks, however, are fairly common in clinical setting (e.g. annotation of clinical interviews, assignment of ICD-9/10 codes to patient records). To address this limitation, in this paper we propose contextual and semantic features and present the results of an extensive experimental evaluation of state-of-the-art supervised machine learning methods in conjunction with lexical and the proposed features for the task of automatic annotation of utterances in clinical interview transcripts with the codebooks consisting of large number of classes. Our study provides a guideline for clinical informatics researchers and practitioners, who consider an option of using machine learning methods for automatic annotation of clinical text in their projects.

Annotation of clinical interview transcripts to distinguish different behavior types is an important part of clinical research aimed at designing effective interventions for many conditions and disorders. In this paper, we focus on the transcripts of Motivational Interviews with obese adolescents (teens) and their caregivers. Childhood obesity is a serious public health concern in the United States and worldwide. Recent estimates indicate that approximately one third (31.8\%) of US children age 2-19 years are overweight and 16.9\% are obese \citep{01}. Adolescents who are obese are likely continue to be obese in adulthood and have a greater risk of heart disease, type 2 diabetes, stroke, cancer, and osteoarthritis \citep{02}. Therefore, there is a need for informatics-based methods to facilitate development of effective interventions for childhood obesity. One approach to designing intervention is Motivational Interviewing (MI), an evidence-based counseling technique to increase intrinsic motivation and self-efficacy for health-related behavior change \citep{03,04}. The goal of a MI counseling session is to encourage patients to explore their own desires, ability, reasons, need for and commitment to the targeted behavior change. These statements, referred to as ``change talk'' (or CT), consistently predict actual behavior change \citep{05} that can be sustained for as long as 34 months after an interview \citep{06}. The process of establishing causal linkages to identify effective communication strategies for eliciting change talk and commitment language in MI involves a resource-intensive qualitative coding process. First, clinical interviews are transcribed and then each utterance is manually annotated with a set of codes from a pre-defined codebook designating specific behavior types. Training human coders to reliably and accurately assign codes to textual fragments requires a large investment of manpower, time and money. For example, in a recent MI study \citep{07}, training coders to reliability took about four months and, once trained, coders required five hours to code every recorded hour. A similar study reported requiring 60 hours of training over six weeks to attain coder reliability, and the actual coding involved two coding passes and six coders \citep{08}.

Automatic annotation of patient utterances in clinical communication is a challenging task, since patients usually come from a variety of cultural and educational backgrounds and
their language use can be quite different \citep{09}. This problem is exacerbated when the interviews are conducted with children and adolescents due to their tendency to use
incomplete sentences and frequently change subjects.

Previous quantitative studies of clinical conversation have resulted in creation of Generalized Medical Interaction Analysis System (GMIAS) \citep{10}, which uses a codebook with generic hierarchical categories. The small-size codebook in Comprehensive Analysis of the Structure of Encounters System (CASES) \citep{11} was designed to annotate several
meta-discursive aspects of medical interviews, such as assigning ``ownership'' of topics and partitioning them into distinct segments (speech acts). It was also shown that the
fragments of transcripts of routine outpatient visits consisting of several speech acts coded using GMIAS and CASES can be annotated as ``information giving'' and ``requesting
information'' \citep{12}. Other related previous studies focused on categorizing assertions of medical problems in clinical narrative into 5 classes (present, absent, possible,
hypothetical, conditional and associated with someone else) using SVM \citep{13} and annotating the utterances in hemodialysis phone dialogue with 3 categories using AdaBoost
classifier \citep{14}. The present study reports the results of comprehensive evaluation of 8 state-of-the-art classifiers (Na\"{i}ve Bayes \citep{15,16,17}, Support Vector Machine
\citep{18,19}, Conditional Random Fields \citep{21,22}, J48 \citep{23}, AdaBoost \citep{24}, Random Forest \citep{42}, DiscLDA \citep{25} and Convolutional Neural Network
\citep{38}) for the task of annotating clinical interviews with a codebook, consisting of a large number of classes. We also propose and experimentally evaluate two novel features
for this task: contextual features based on the label of the preceding textual fragment and semantic features based on the distribution of words in the annotated fragment over a linguistic lexicon.

Codes automatically assigned to transcripts using classification models presented above allow to abstract away from the raw text  of the interviews and unequivocally establish the casual relationship between provider communication and client change talk (CT). Two recent studies \citep{46,47} have examined causality in treatment communication sequences using sequential analysis techniques to assess the probability that CT actually follows specific provider behaviors. Both studies found that provider use of MICO communication behaviors was followed by CT more frequently than expected and followed by client counter-CT (CCT; client statements against making changes) less frequently than expected by chance. Again, while suggestive of important PPC sequences, these studies were conducted among adults in treatment for alcohol abuse. In this project, we used a probabilistic model, in particular, first order markov model for the analysis of sequential data to determine provider-patient communication sequences that are likely to translate into change talk and commitment language.

\begin{table}[htb!]
\caption{\textbf{Distribution of utterances over 16 classes in the caregiver dataset.}}    
\label{table:01}
\centering
\begin{tabular}{lp{7cm}rr}
\hline
\hline
Code  & Description & Utterance & \% \\
\hline
209 &	Caregiver Change Talk, negative &	297 & 6.82 \\
212 &	Caregiver Change Talk, positive &	1107 & 25.40 \\
232 &	Low Uptake, positive &	518 & 11.89 \\
235 &	High uptake &	231 & 5.30 \\
301 &	Structure Session &	206 & 4.73 \\
302 &	General Information, positive &	309 & 7.09 \\
305 &	Emphasize Autonomy &	148 & 3.40 \\
306 &	Closed question, Elicit Feedback &	50 & 1.15 \\
307 &	Support &	108 & 2.48 \\
308 &	Affirm &	289 & 6.63 \\
315 &	Reflect, change talk positive, about caregiver &	659 & 15.12 \\
329 &	Self-disclose &	44 & 1.01 \\
330 &	Statement, other &	121 & 2.78 \\
331 &	Open-ended question, elicit change talk positive &	200 & 4.59 \\
343 &	Open-ended question, target behavior neutral &	33 & 0.76 \\
344 &	Open-ended question, elicit barriers &	38 & 0.87 \\
\hline
\hline
\end{tabular}
\end{table}

\begin{table}[htb!]
\caption{\textbf{Distribution of sequences over the adolescent interview session}} 
\label{table:0001}
\centering
\begin{tabular}{lrr}
\hline
\hline
Sequence type  & \# of Sequences & \% \\
\hline
Successful & 880 & 82.08\% \\
Unsuccessful & 192 & 17.92\% \\
Total & 1072 & 100\% \\
\hline
\hline
\end{tabular}
\end{table}

\section*{Methods}
\subsection*{\textit{Data collection}}
The golden standard for evaluation of machine learning methods was created based on the transcripts of motivational interviews conducted by the clinicians at the Pediatric
Prevention Research Center (PPRC) of Wayne State University. Each interview is comprised of two parts: conversation of a clinician with an adolescent followed by a conversation of
a clinician with the adolescent's caregiver. All adolescents in this study were between the ages of 12 and 17 (M = 14.7, SD = 1.63) and most were female (n = 27). Most caregivers
were biological mothers (n = 36), who were married or living with a partner (n = 25). The median family income was \$16,000--\$21,999 ranging from less than \$1,000 to
\$50,000--\$74,999. The audio recordings of the interviews were first transcribed and segmented into utterances belonging to adolescents, caregivers, and counselors, preserving the
sequence of utterances. Transcripts were then manually annotated by trained human coders according to MYSCOPE \citep{07}, a specialized codebook including a large number of
behavior codes, which was developed by an interdisciplinary team including a clinical psychologist, a nutrition scientist, a communication scientist, a linguist and a community
health worker specifically for annotating motivational interviews with obese adolescents. A primary coder independently coded interview sessions and a secondary coder co-coded a
randomly selected 20\% of the transcripts to monitor reliability ($\kappa$ = 0.696) \citep{07}. The MYSCOPE codebook contains a total of 115 different codes that are grouped into
the youth, caregiver, and counselor code groups. The experimental datasets for this work were constructed based on the transcripts of 37 motivational interview sessions, which
include a total of 11,353 segmented and annotated utterances. These utterances have been further partitioned into two subsets based on the structure of motivational interview
sessions: one dataset that includes all utterances from the adolescent sessions (6,579 samples) and the other dataset that includes all utterances from the caregiver sessions
(4,774 samples). A fragment of an adolescent session transcript is presented in Table~\ref{table:06}.

@@

To conduct a detailed analysis of performance of classification methods, we used the following two-stage process to create the codebooks with different number of codes for
adolescent and caregiver sessions. In the first stage, we merged conceptually similar behavior codes as well as the codes with similar data distributions, while in the second stage, we eliminated the codes with insufficient data samples. In case of the adolescent sessions, we started with 55 adolescent session-specific codes and, after merging the codes with subtle differences (e.g. converting valiances of change talk CHT+1, CHT+2 and CHT+3 into CHT+), obtained a codebook with 41 classes. We further reduced this codebook to 20 classes after merging 21 classes with similar sample distributions. After eliminating the codes that had less than 10 data samples (to ensure that there can be at least one sample of each class in each fold when using 10-fold cross validation experimental design), we obtained a third codebook with 17 codes. Using the same approach, we created the codebooks containing 58, 19, and 16 caregiver session-specific codes. Table~\ref{table:01} shows the distribution of utterances over 16 classes in the caregiver session transcripts. As follows from Table~\ref{table:01}, the distribution of utterances over classes is highly imbalanced even for the codebook of the smallest size, which is fairly common for clinical text. To create the gold standard for our second experiment, we partitioned the adolescent transcripts into sequence of successful and unsuccessful codes which are ended with either negative or positive change talk and commitment language. Table~\ref{table:0001} depicts the distribution of PPC code sequences over the adolescent dataset.

After creating the codebooks, we pre-processed the dataset using the Snowball stemmer available as part of the Weka \citep{26} machine learning toolkit\footnote{
http://www.cs.waikato.ac.nz/ml/weka/}. We also found out that stopword removal decreased the performance of classification models for our task (e.g., in case of the codebook
consisting of 17 classes, the accuracy of Na\"{i}ve Bayes decreased from 67\% to 47.10\%, while the accuracy of SVM decreased from 70.76\% to 55.26\%). A likely reason is that,
although negations are typically considered as stopwords, they are fairly important clues for inferring certain behavior types (e.g., removing the stopword ``not'' completely
transforms the meaning of a phrase ``not great'').

@@

\subsection*{\textit{Performance of Markov model in conjunction with determining PPC code sequences that are most likely translate into CT}}
Since clinical researchers tried to increase the desire and ability of adolescent to change their current behavior to target behavior, we used sequence of successful and unsuccessful codes only from adolescent session for our sequential analysis. Summary of performance for the classification of sequential data is presented in Table~\ref{table:sequence}. 
From Table~\ref{table:sequence}, it follows that our designed markov model works very well in terms of precision and achieves precision 0.7574 with F1-measure 0.7092. The strength of the model for the classification of each type of sequences is illustrated by providing the confusion matrix in Table~\ref{table:conf}. It also shows that successful patterns are identified more accurately compare to unsuccessful motivational interview sequences due to the imbalance of data.

\begin{table}
\parbox{.45\linewidth}{
\centering
\caption{\textbf{Performance of markov model for the classification of PPC code sequences}}
\label{table:sequence}
\begin{tabular}{ll}
\hline
\hline
Metrics & Value \\
\hline
Accuracy &	0.6803 \\
Precision &	0.7574 \\
Recall & 0.6803 \\
F1-Measure & 0.7092 \\
\hline
\hline
\end{tabular}
}
\hfill
\parbox{.45\linewidth}{
\centering
\caption{\textbf{Confusion matrix of proposed markov model for the classification of sequence of codes}}
\label{table:conf}
\begin{tabular}{llll}
\hline
\hline
& Successful & Fail & Total\\
\hline
Successful & 642 & 242 & 884 \\
Fail & 102 & 90 & 192 \\
Total & 744 & 332 & 1076 \\
\hline
\hline
\end{tabular}
}
\end{table}

@@

\begin{table}[htb!]
\caption{\textbf{Top 5 most likely successful and unsuccessful interview sessions}}    
\label{table:topsec}
\centering
\begin{tabular}{lll}
\hline
\hline
No. & Successful Sequences &  Unsuccessful Sequences\\
\hline
1 & 311\rightarrow112\rightarrow301\rightarrow500 & 343\rightarrow109\rightarrow312\rightarrow400 \\
2 & 310\rightarrow112\rightarrow311\rightarrow500 & 311\rightarrow109\rightarrow312\rightarrow400 \\
3 & 311\rightarrow117\rightarrow309\rightarrow500 & 327\rightarrow120\rightarrow312\rightarrow400 \\
4 & 312\rightarrow112\rightarrow311\rightarrow500 & 314\rightarrow103\rightarrow314\rightarrow400 \\
5 & 331\rightarrow112\rightarrow311\rightarrow500 & 344\rightarrow109\rightarrow326\rightarrow400 \\
\hline
\hline
\end{tabular}
\end{table}

>>>>

\title{Designing an effective intervention of motivational interview using sequence classification methods}

@@

\noindent{\bf Abstract}
\textit{Coming soon!}

\section*{Introduction}

In the recent times, more and more data from various domains being produced in the form of event sequences. Thus, sequence classification has become an important task in sequence mining. Sequences can contain discrete (e.g., symbolic sequences such as text, proteins, or DNA) or continuous events (e..g, time series such as ECGs or stocks), depending on the event types. Usually, sequences don't have explicit features and suffer with high dimension of feature space. Even the sequential nature of features is too difficult to capture, which make sequence classification, a more challenging task than feature-based classifications. The sequence classification methods can be divided into three large categories: feature-based classification, distance based method and model based classifications. In feature-based classification, a sequence is transformed into a single vector of features. A supervised learning algorithm such as an support vector machine \cite{leslie2004fast} or decision tree \cite{chuzhanova1998feature} can be used to train the classifier. An improvement of feature-based classifier such as shapelet \cite{ye2009time} or pattern-based \cite{kudenko1998feature, lesh1999mining} approach is used for sequence classifications. On the other hand, distance based method measure the similarity between sequences to determine the quality of the classification significance. The most commonly used distance function is euclidian distance \cite{keogh2003need}, while Dynamic Time Wrapping \cite{keogh2000scaling} is used when more flexible matching is desired in time series data. The third category of sequence classification is model based classifiers such as Hidden Markov Model \cite{rabiner1989tutorial} (HMM) or other statistical models. Recently, a hierarchical approach \cite{nallam2016effective} is proposed for sequence classification.

Sequence classification has a broad range of real word applications such as information retrieval, genomic analysis, health informatics, finance and abnormal detection. In genomic research, sequence classification is widely used to detect the function of a new protein \cite{deshpande2002evaluation}. In health informatics, ECG can be used as a multi-dimensional time series data to classify an individual as healthy or patient with heart disease \cite{wei2006semi}. Another example is to classify rumour stance in Twitter \cite{lukasik2016hawkes}. In information retrieval, sequence classification is employed to build  models for classifying documents into different topic categories \cite{sebastiani2002machine} such as sports, politics, news, finance and style. Sequence classification is also used in anomaly detection such as abnormal access to systems \cite{lane1999temporal}, malware detection \cite{drew2017polymorphic} and money laundering \cite{liu2008sequence}. Other interesting applications include classifying query log sequences to distinguish web-robots from human users \cite{duskin2009distinguishing, tan2004discovery}. In computer vision, sequence classification is applied for classifying images \cite{li2000image} and videos \cite{zhang2016tensor}.

In this paper, our target area of application is motivational interview. In particular, this paper describes our work on the classification of utterance sequences. Classification of utterance sequences to distinguish different behavior outcomes such as positive change talk or commitment verses negative change talk or commitment, is an important part of clinical research aimed at designing effective interventions for many conditions and disorders. In this study, we focus on the transcripts of Motivational Interviews with obese adolescents (teens) and their caregivers. Childhood obesity is a serious public health concern in the United States and worldwide. Recent estimates indicate that approximately one third (31.8\%) of US children age 2-19 years are overweight and 16.9\% are obese \citep{01}. Adolescents who are obese are likely continue to be obese in adulthood and have a greater risk of heart disease, type 2 diabetes, stroke, cancer, and osteoarthritis \citep{02}. Therefore, there is a need for informatics-based methods to facilitate development of effective interventions for childhood obesity. One approach to designing intervention is Motivational Interviewing (MI), an evidence-based counseling technique to increase intrinsic motivation and self-efficacy for health-related behavior change \citep{03,04}. The goal of a MI counseling session is to encourage patients to explore their own desires, ability, reasons, need for and commitment to the targeted behavior change. These statements, referred to as ``change talk'' (or CT), consistently predict actual behavior change \citep{05} that can be sustained for as long as 34 months after an interview \citep{06}.

Our previous studies \cite{kotov2015interpretable, hasan2016study} explored several machine learning methods for the automatic annotation of clinical interview fragments, which uses a codebook \cite{carcone2013provider} with 115 different behavior codes. In this study, probabilistic models are employed for the analysis of sequential data to determine provider-patient communication sequences that are likely to translate into change talk and commitment language. Currently, a very similar work \cite{jaber2016multi} was done on education data for sequence classifications. Although Hidden Markov Model was originally applied to speech recognition \cite{rabiner1989tutorial}, this is one of the most successful method \cite{mutsam2016maximum, eickeler1998hidden, won2004training, chai2001folk} for sequence classification. Therefore, we utilize Markov Chain and Hidden Markov Model for the classification of utterance sequences.

\section*{Methods}
\subsection*{\textit{Data collection}}
The MYSCOPE \citep{07} codebook contains a total of 115 different codes that are grouped into the youth, caregiver, and counselor code groups. The experimental datasets for this work were constructed based on the transcripts of 37 motivational interview sessions, which
include a total of 11,353 segmented and annotated utterances. These utterances have been further partitioned into two subsets based on the structure of motivational interview
sessions: one dataset that includes all utterances from the adolescent sessions (6,579 samples) and the other dataset that includes all utterances from the caregiver sessions
(4,774 samples). A fragment of an adolescent session transcript is presented in Table~\ref{table:06}.

@@

To create the gold standard for our second experiment, we partitioned the adolescent transcripts into sequence of successful and unsuccessful codes which are ended with either negative or positive change talk and commitment language. Table~\ref{table:0001} depicts the distribution of PPC code sequences over the adolescent dataset.

@@

\subsection*{\textit{Performance of Markov model in conjunction with determining PPC code sequences that are most likely translate into CT}}
Since clinical researchers tried to increase the desire and ability of adolescent to change their current behavior to target behavior, we used sequence of successful and unsuccessful codes only from adolescent session for our sequential analysis. Summary of performance for the classification of sequential data is presented in Table~\ref{table:sequence}. 
From Table~\ref{table:sequence}, it follows that our designed markov model works very well in terms of precision and achieves precision 0.7574 with F1-measure 0.7092. The strength of the model for the classification of each type of sequences is illustrated by providing the confusion matrix in Table~\ref{table:conf}. It also shows that successful patterns are identified more accurately compare to unsuccessful motivational interview sequences due to the imbalance of data.\\

\begin{table}[h]
\centering
\caption{Performance of markov models for the classification of normal PPC code sequence.}
\label{tab:result_norm_seq}
  \begin{tabular}{|l|l|l|l|l|l|}
  \hline
   \textbf{Model} & \textbf{Order}  & \textbf{Precision}  & \textbf{Recall} & \textbf{F-Measure} & \textbf{Accuracy}\\ \hline

 \multirow{2}{*}{General Markov Chain} & First order & 0.7387 & 0.7686 & 0.7532 & 0.7686\\\cline{2-6}
 & Second order & 0.6889 & 0.7889 & 0.7352 & 0.7889\\ \hline
 \multirow{2}{*}{Hidden Markov Model} & First order & \textbf{0.7980} & 0.8059 & \textbf{0.7989} & 0.8059\\ \cline{2-6}
 & Second order & 0.7400 & \textbf{0.8449} & 0.7822  & \textbf{0.8449}\\ \hline

  \end{tabular}
\end{table}

\begin{table}[h]
\centering
\caption{Performance of markov models for the classification of alternate PPC code sequence.}
\label{tab:result_alt_seq}
  \begin{tabular}{|l|l|l|l|l|l|}
  \hline
   \textbf{Model} & \textbf{Order}  & \textbf{Precision}  & \textbf{Recall} & \textbf{F-Measure} & \textbf{Accuracy}\\ \hline

 \multirow{2}{*}{General Markov Chain} & First order & 0.9777 & 0.9437 & 0.9604 & 0.9437\\\cline{2-6}
 & Second order & \textbf{0.9778} & 0.9694 & \textbf{0.9736} & 0.9694\\ \hline
 \multirow{2}{*}{Hidden Markov Model} & First order & 0.9392 & 0.9313 & 0.9253 & 0.9392\\ \cline{2-6}
 & Second order & 0.9704 & \textbf{0.9713} & 0.9699  & \textbf{0.9713}\\ \hline

  \end{tabular}
\end{table}

@@
